---
title: "Assignment2_Data622"
author: "Mubashira Qari"
date: "2025-10-04"
output: html_document
---
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
  pdf_document:
    toc: true
---
# Bank Marketing Dataset Variables  

The dataset’s main purpose is to predict whether a bank client will subscribe to a term deposit (`y = yes/no`) after a marketing campaign. Each variable provides information that may help explain or predict that outcome.  

---

## Client Information (Who the Customer Is)  

- **age (numeric)**  
  - *Recording:* Client’s age in years.  
  - *Purpose:* Age often influences financial decisions (e.g., younger clients may avoid long-term deposits, older clients may be more interested in safe investments).  

- **job (categorical)**  
  - *Recording:* Type of job (e.g., admin, blue-collar, technician).  
  - *Purpose:* Occupation reflects income stability and financial behavior.  

- **marital (categorical)**  
  - *Recording:* Marital status (single, married, divorced/widowed).  
  - *Purpose:* Family responsibilities affect savings and investment behavior.  

- **education (categorical)**  
  - *Recording:* Highest education level completed.  
  - *Purpose:* Education level may indicate financial literacy and likelihood to invest.  

- **default (categorical)**  
  - *Recording:* Whether client has credit in default ("yes", "no", "unknown").  
  - *Purpose:* Default history reflects financial risk and trustworthiness.  

- **housing (categorical)**  
  - *Recording:* Whether client has a housing loan.  
  - *Purpose:* Mortgage holders may have less disposable income for deposits.  

- **loan (categorical)**  
  - *Recording:* Whether client has a personal loan.  
  - *Purpose:* Personal loans indicate financial obligations that may reduce likelihood of subscribing.  

---

## Last Contact Information (How They Were Reached)  

- **contact (categorical)**  
  - *Recording:* Communication type ("cellular" or "telephone").  
  - *Purpose:* Some contact methods are more effective than others.  

- **month (categorical)**  
  - *Recording:* Month of the last contact.  
  - *Purpose:* Seasonal effects — campaign success may vary across the year.  

- **day_of_week (categorical)**  
  - *Recording:* Day of the week client was contacted.  
  - *Purpose:* Some days may be better for reaching clients (e.g., midweek vs. Monday).  

- **duration (numeric)**  
  - *Recording:* Call duration in seconds.  
  - *Purpose:* Longer conversations often mean higher engagement.  
  - *Note:* Duration is only known after the call, so it cannot be used in real prediction models.  

---

## Campaign History (Past Interactions)  

- **campaign (numeric)**  
  - *Recording:* Number of contacts during this campaign (including the last one).  
  - *Purpose:* Too many contacts may annoy clients, reducing success.  

- **pdays (numeric)**  
  - *Recording:* Days since last contact from a previous campaign (999 = never contacted).  
  - *Purpose:* Recency matters; recently contacted clients may behave differently.  

- **previous (numeric)**  
  - *Recording:* Number of contacts before this campaign.  
  - *Purpose:* Reflects persistence of bank marketing; may affect likelihood of success.  

- **poutcome (categorical)**  
  - *Recording:* Outcome of the previous campaign ("success", "failure", "nonexistent").  
  - *Purpose:* Past behavior often predicts future responses.  

---

## Economic Context (Overall Environment)  

- **emp.var.rate (numeric)**  
  - *Recording:* Employment variation rate (quarterly).  
  - *Purpose:* Measures labor market changes — people may invest more when jobs are stable.  

- **cons.price.idx (numeric)**  
  - *Recording:* Consumer Price Index (monthly).  
  - *Purpose:* Inflation affects purchasing power and saving behavior.  

- **cons.conf.idx (numeric)**  
  - *Recording:* Consumer Confidence Index (monthly).  
  - *Purpose:* High confidence = more willingness to invest, low confidence = caution.  

- **euribor3m (numeric)**  
  - *Recording:* Euribor 3-month interest rate.  
  - *Purpose:* Competes with deposit rates — higher Euribor may reduce deposit attractiveness.  

- **nr.employed (numeric)**  
  - *Recording:* Number of employees (quarterly labor market indicator).  
  - *Purpose:* Reflects macroeconomic health; higher employment often correlates with more savings.  

---

## Target Variable  

- **y (binary: "yes" / "no")**  
  - *Recording:* Whether the client subscribed to a term deposit.  
  - *Purpose:* This is the outcome to predict.  

---

## PART I: Exploratory Data Analysis (EDA) {.tabset}



```{r warning=FALSE}
# Load Libraries

library(dplyr)
library(tidyverse)
library(psych)
library(ggplot2)
library(plotly)
library(tidyr)
library(corrplot)
library(ggpubr)
library(naniar)     # for missing value visualization
library(DataExplorer) # optional: automated EDA
library(forcats)
library(caret)
library(recipes)
library(themis)
library(smotefamily)


```



```{r warning=FALSE}
# Load Dataset

url <- "https://raw.githubusercontent.com/uzmabb182/Data_622/refs/heads/main/Assignment_1_EDA/bank-additional-full.csv"
bank_additional_df <- read.csv2(url, stringsAsFactors = FALSE)
#head(bank_additional_df)
bank_additional_df
```



```{r}
# Basic structure
str(bank_additional_df)

```


```{r}
# Dimensions
dim(bank_additional_df)   # rows, columns
nrow(bank_additional_df)  # number of rows
ncol(bank_additional_df)  # number of columns

```


```{r}
# Column names
names(bank_additional_df)

```


```{r}
# Summary statistics for all variables
summary(bank_additional_df)

```
Dataset Summary:
The dataset consists of 41,188 observations and 21 variables describing bank clients, marketing campaign details, and macroeconomic indicators.
Most variables are categorical, with several containing “unknown” values that need preprocessing.
Economic indicators are incorrectly stored as text and should be converted to numeric.
The duration feature should be excluded from predictive modeling due to data leakage.
The target variable y is binary (“yes”/“no”), confirming this as a classification problem.

## Are there any missing values and how significant are they? 

```{r}
# First and last few records
head(bank_additional_df, 10)
tail(bank_additional_df, 10)

# Missing values per column
missing_summary <- bank_additional_df %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Missing_Count") %>%
  mutate(Missing_Percent = round(Missing_Count / nrow(bank_additional_df) * 100, 2)) %>%
  arrange(desc(Missing_Count))

missing_summary

```

## Find Unwanted Columns

```{r}

# Identify Columns with a Single Unique Value

sapply(bank_additional_df, function(x) length(unique(x)))

```
```{r}
# Detect Columns with Mostly “Unknown” or Missing Values
# Count 'unknown' or NA values per column
sapply(bank_additional_df, function(x) sum(x == "unknown" | is.na(x)))

```

```{r}
# Check for Duplicated or Irrelevant Columns
duplicated_cols <- names(bank_additional_df)[duplicated(as.list(bank_additional_df))]
duplicated_cols

```
Conclusion:

The dataset has 41,188 records and 21 features — large enough for robust machine learning modeling.

The target variable is y with two possible outcomes — “yes” or “no”, making this a binary classification problem.

Most of your columns are categorical (character), not numeric — you’ll need to encode them (e.g., one-hot encoding or dummy variables) before modeling.

A few columns that should be numeric (emp.var.rate, cons.price.idx, cons.conf.idx, euribor3m, nr.employed) are currently stored as characters (chr).
These must be converted to numeric for analysis or correlation checks.

The column duration represents call length, which is not known before prediction time → you should exclude it to avoid data leakage.

Some columns (like default, housing, loan) include “unknown” values — these must be handled (replace, impute, or treat as a separate category).

Out of the 21 variables in the Bank Marketing dataset, six contain “unknown” or missing values — specifically job, marital, education, default, housing, and loan.

Among these, the default column has the highest missing rate (~20%) and may be excluded or handled cautiously.

The other variables have relatively few missing values, so “unknown” can be treated as a valid category or imputed with the most frequent value.

The remaining 15 variables are complete and ready for analysis.

```{r}
# Watch for Data Leakage
# duration measures the call length, but you only know it after the call is made — which means it leaks information from the future into the model.
#bank_additional_df <- subset(bank_additional_df, select = -c(duration))

```

```{r}
# analyze the relationship between all categorical variables and your label (y)
# Make sure "unknown" values are converted to proper NA, and your target variable y is a factor.

# Replace "unknown" with NA
bank_additional_df[bank_additional_df == "unknown"] <- NA

# Convert target variable to factor
bank_additional_df$y <- as.factor(bank_additional_df$y)
bank_additional_df
```

```{r}
# Identify All Categorical Variables

cat_vars <- names(bank_additional_df)[sapply(bank_additional_df, function(x) is.character(x) | is.factor(x))]
cat_vars <- setdiff(cat_vars, "y")  # remove target variable
cat_vars

```
```{r}
#Run Chi-Square Test for Each Variable vs y

# This test checks if the categorical feature and the target variable are statistically related.

# Chi-square test for each categorical variable vs y
chi_results <- sapply(cat_vars, function(var) {
  tbl <- table(bank_additional_df[[var]], bank_additional_df$y)
  test <- chisq.test(tbl)
  p_value <- round(test$p.value, 6)
  return(p_value)
})

# View results
chi_results <- sort(chi_results)
chi_results

```

```{r}
# Visualize Relationships (Proportion Bar Charts)

library(ggplot2)

for (var in cat_vars) {
  print(
    ggplot(bank_additional_df, aes_string(x = var, fill = "y")) +
      geom_bar(position = "fill") +
      labs(
        title = paste("Relationship Between", var, "and Term Deposit (y)"),
        y = "Proportion of Responses",
        x = var
      ) +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
  )
}

```

```{r}
sig_vars <- names(chi_results[chi_results < 0.05])

summary_df <- data.frame(
  Variable = names(chi_results),
  P_Value = chi_results,
  Significant = ifelse(chi_results < 0.05, "Yes", "No")
)

summary_df

```
```{r}
library(ggplot2)

ggplot(summary_df, aes(x = reorder(Variable, -log10(P_Value)), y = -log10(P_Value))) +
  geom_col(fill = "steelblue") +
  geom_hline(yintercept = -log10(0.05), color = "red", linetype = "dashed") +
  labs(
    title = "Strength of Association Between Categorical Variables and Target (y)",
    x = "Variable",
    y = "-log10(p-value)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
```{r}
# Identify Numeric Columns
# Identify numeric columns
num_vars <- names(bank_additional_df)[sapply(bank_additional_df, is.numeric)]
num_vars

```
```{r}
# Summary Statistics by Label (y)Summary Statistics by Label (y)
library(dplyr)

bank_additional_df %>%
  group_by(y) %>%
  summarise(across(all_of(num_vars), list(mean = mean, sd = sd), na.rm = TRUE))

```
```{r}
# Statistical Relationship: t-test or Wilcoxon Test
# test if each numeric variable is significantly different between “yes” and “no” clients.
num_tests <- sapply(num_vars, function(var) {
  ttest <- t.test(bank_additional_df[[var]] ~ bank_additional_df$y)
  return(round(ttest$p.value, 5))
})

num_tests <- sort(num_tests)
num_tests

```
```{r}
# Compare numeric variable distributions for “yes” and “no” - Bar Plot

library(ggplot2)

for (var in num_vars) {
  print(
    ggplot(bank_additional_df, aes(x = y, y = .data[[var]], fill = y)) +
      geom_boxplot() +
      labs(
        title = paste("Distribution of", var, "by Subscription (y)"),
        x = "Subscription (y)",
        y = var
      ) +
      theme_minimal()
  )
}

```

```{r}

# Density Plot

for (var in num_vars) {
  print(
    ggplot(bank_additional_df, aes(x = .data[[var]], color = y, fill = y)) +
      geom_density(alpha = 0.4) +
      labs(
        title = paste("Density Plot of", var, "by Subscription (y)"),
        x = var,
        y = "Density"
      ) +
      theme_minimal()
  )
}

```

```{r}
# encode the target (y) as numeric (no = 0, yes = 1) and calculate correlation with numeric variables.

bank_additional_df$y_num <- ifelse(bank_additional_df$y == "yes", 1, 0)

correlations <- sapply(num_vars, function(v) cor(bank_additional_df[[v]], bank_additional_df$y_num, use = "complete.obs"))
correlations <- sort(correlations, decreasing = TRUE)
correlations

```
```{r}
# Combine Statistical and Visual Findings
summary_df <- data.frame(
  Variable = names(num_tests),
  P_Value = round(num_tests, 5),
  Correlation = round(correlations[names(num_tests)], 3)
)

summary_df$Significant <- ifelse(summary_df$P_Value < 0.05, "Yes", "No")
summary_df

```
Interpretation summary:

Conclusion:
All numeric variables in the Bank Marketing dataset show statistically significant relationships with the target variable (y).
Moderate correlations are found for duration (0.405), nr.employed (-0.355), pdays (-0.325), euribor3m (-0.308), and emp.var.rate (-0.298).
These results suggest that both macroeconomic factors and client interaction history significantly influence subscription likelihood.
Lower interest rates, fewer employees (economic downturns), and recent contact history increase the chance of a client subscribing.
The duration variable shows a strong positive association but represents post-call information, making it unsuitable for model training due to data leakage.
Weak correlations for age and consumer confidence index imply these variables have limited predictive value.