---
title: "final_project_data_622"
author: "Mubashira Qari"
date: "2025-12-07"
output: html_document
---

Dataset: Phishing_Email.csv

Goal: classify emails as phishing vs legitimate using text

Chapter 1–10 method: Naive Bayes (Chapter 4)

Chapter 11–15 method: Random Forest (Chapter 12 – ensemble learning)


STEP 1 – Install and load required packages

```{r message=FALSE, warning=FALSE}
#install.packages("tm")          # text mining
#install.packages("e1071")       # Naive Bayes
#install.packages("randomForest")# Random Forest
#install.packages("ggplot2")     # simple plots (EDA)
library(tm)
library(e1071)
library(randomForest)
library(ggplot2)

```

Explanation:

tm helps with cleaning and turning raw text into a document-term matrix.

e1071 contains the naiveBayes() function.

randomForest contains the randomForest() function.

ggplot2 is for visualizations.

STEP 2 – Load the dataset and inspect it

```{r warning=FALSE}
# Load Dataset

url <- "https://raw.githubusercontent.com/uzmabb182/Data_622/refs/heads/main/final_project_data_622/Phishing_Email.csv"
phishing_email_df <- read.csv(url, stringsAsFactors = FALSE)
head(phishing_email_df)

```



```{r}
# Look at structure and first few rows
str(phishing_email_df)
head(phishing_email_df)
names(phishing_email_df)

```

Email.Text → the email body (unstructured text)

Email.Type → the label: "Safe Email" or "Phishing Email"

STEP 3 – Create a clean Label column for modeling

Right now, your label is in Email.Type as text like "Safe Email" and "Phishing Email".
turn that into a factor called Label with two values: "legitimate" and "phishing".

3.1 – First, inspect the values in Email.Type

```{r}
table(phishing_email_df$Email.Type)

```

3.2 – Create the Label factor column

```{r}
phishing_email_df$Label <- factor(
  phishing_email_df$Email.Type,
  levels = c("Safe Email", "Phishing Email"),
  labels = c("legitimate", "phishing")
)

# Check the new Label column
table(phishing_email_df$Label)
str(phishing_email_df$Label)

```

Explanation

phishing_email_df$Email.Type = original labels ("Safe Email" / "Phishing Email").

factor(..., levels = ..., labels = ...)

levels = what is currently in the data.

labels = how we want to rename them for modeling.

We create a new column Label (we’re not overwriting Email.Type, we keep it as reference).

STEP 4 – Create and inspect the text corpus

We’ll now turn the Email.Text column (raw email content) into a corpus, which is just a collection of text documents used for text mining.

4.1 – Create the corpus

```{r}
library(tm)  # make sure this is loaded

corpus <- VCorpus(VectorSource(phishing_email_df$Email.Text))

```

Explanation

VectorSource(phishing_email_df$Email.Text)
This tells R: “treat each row of Email.Text as a separate document.”

VCorpus(...)
Creates a volatile corpus in memory, which tm can clean and transform.

4.2 – Inspect a few emails from the corpus

```{r}
# Look at the first email
corpus[[1]]$content

# Look at, say, the 10th email
corpus[[10]]$content

```

Explanation

corpus[[1]]$content displays the raw text of the first email.

This helps you see what the model will be working with before cleaning.

STEP 5 – Clean the Text Corpus

We apply a series of transformations to standardize the email text.

5.1 – Convert everything to lowercase


```{r}
corpus_clean <- tm_map(corpus, content_transformer(tolower))

```

Explanation:
Machine learning treats “Free”, “FREE”, and “free” as different words unless we normalize them. 
Lowercasing avoids duplication.

5.2 – Remove numbers

```{r}
corpus_clean <- tm_map(corpus_clean, removeNumbers)

```

Many numbers do not contribute to detecting phishing (unless you're doing more complex NLP).

5.3 – Remove punctuation

```{r}
corpus_clean <- tm_map(corpus_clean, removePunctuation)

```

Punctuation does not carry useful meaning for most traditional ML models.

5.4 – Remove common English stopwords

```{r}
corpus_clean <- tm_map(corpus_clean, removeWords, stopwords("en"))

```

Stopwords are very common words like:
the, and, or, is, to, be, of, at
They appear so often they do not help classification.

5.5 – Remove extra whitespace

```{r}
corpus_clean <- tm_map(corpus_clean, stripWhitespace)

```

This cleans up spaces left behind after removing stopwords and punctuation.

5.6 – Inspect cleaned text

```{r}
corpus_clean[[1]]$content

```

now we see a simplified version of the original email.

STEP 6 — Create the Document-Term Matrix (DTM)

This step converts your cleaned text into numbers (word counts), which allows machine learning models to use the data.

6.1 – Build the DTM

```{r}
dtm <- DocumentTermMatrix(corpus_clean)
dtm

```

Explanation

Each row = one email

Each column = one unique word

Each cell = number of times the word appears in that email

STEP 7 — Reduce sparsity (remove extremely rare words)

Rare words clutter the model and slow everything down.
We keep words that appear in at least 1% of emails.

```{r}
dtm_sparse <- removeSparseTerms(dtm, 0.99)
dtm_sparse

```

Explanation

0.99 keeps terms appearing in at least 1% of emails

This usually reduces terms from 100,000+ to a few hundred or thousand

STEP 8 — Convert the DTM to a Data Frame and Add Label

8.1 — Convert DTM to a matrix and then to a data frame

```{r}
emails_words <- as.data.frame(as.matrix(dtm_sparse))
str(emails_words)

```
Explanation

The machine learning algorithms (Naive Bayes, Random Forest) require a standard R data frame, not a DocumentTermMatrix.

This step turns your DTM into rows (emails) and columns (word counts).

8.2 — Clean column names

Some words contain characters that are invalid in R column names.

```{r}
emails_words <- setNames(emails_words, make.names(names(emails_words)))

```

Explanation

make.names() turns names like "1-800" into something legal like X1.800.

8.3 — Add the Label column

```{r}
emails_words$Label <- phishing_email_df$Label

str(emails_words$Label)
table(emails_words$Label)

```
Explanation

We copy previously created Label factor (legitimate / phishing) over to the new data frame.

Now each row contains:
email word counts + the correct label

This is the dataset we train models on.
