---
title: "Final_Project_Data622"
author: "Mubashira Qari"
date: "2025-11-22"
output:
 html_document:
    toc: true
    toc_float: true
    number_sections: true
 pdf_document:
    toc: true
---
Choice of methodologies:

Chapter 1–10 method: Decision Tree (CART / rpart) — interpretable, good baseline for classification problems.

Chapter 11–15 method: Gradient Boosting (xgboost / gbm) — a powerful ensemble boosting method to improve predictive accuracy.

Problem statement (short)
Predict whether an e-commerce order will be returned (returned column: Yes/No) using order, product, price, shipping, and customer features. 
Business objective: reduce returns by identifying high-risk orders so interventions (quality checks, different shipping, better product descriptions) can be targeted.

1. Setup — packages and seed

```{r message=FALSE, warning=FALSE}

library(tidyverse)
library(lubridate)
library(caret)
library(rpart)
library(rpart.plot)
library(xgboost)
library(pROC)
library(recipes)
library(yardstick)
library(vip)        # variable importance for xgboost
set.seed(1234)

```

Explanation: load tidy tools for data prep, caret for modeling workflow and resampling, rpart for decision trees, xgboost for boosting, pROC for ROC/AUC evaluation. 
set.seed() ensures reproducible splits and resampling.

```{r warning=FALSE}
# Load Dataset

url <- "https://raw.githubusercontent.com/uzmabb182/Data_622/refs/heads/main/final_project_data_622/ecommerce_sales.csv"
ecom_df <- read.csv(url, stringsAsFactors = FALSE)
head(ecom_df)

```
2. Problem Definition

```{r}
# Business problem: reduce returns by predicting "returned" = "Yes" or "No"
# Data science problem: binary classification (returned vs not returned)

```

Explanation: map business problem to binary classification.

3. Data cleaning & feature engineering

```{r}
# assume ecom_df is already in the workspace
# initial glance
glimpse(ecom_df)

# 3.1 Convert types properly (fixes you used earlier)
ecom_df <- ecom_df %>%
  mutate(
    order_date = ymd(order_date),     # dates are like "2023-12-23"
    category = factor(category),
    payment_method = factor(payment_method),
    region = factor(region),
    returned = factor(returned, levels = c("No","Yes")), # ensure No first
    customer_gender = factor(customer_gender)
  ) 

# 3.2 Basic feature engineering
ecom_df <- ecom_df %>%
  mutate(
    order_month = factor(month(order_date, label = TRUE)),
    order_wday = factor(wday(order_date, label = TRUE)),
    unit_price = if_else(quantity > 0, total_amount / quantity, price), 
      # fallback if quantity 0 — guard against divide by zero
    high_value_order = if_else(total_amount > quantile(total_amount, 0.75, na.rm=TRUE), "Yes","No"),
    high_value_order = factor(high_value_order),
    age_group = case_when(
      customer_age < 25 ~ "Under25",
      customer_age >= 25 & customer_age < 45 ~ "25to44",
      customer_age >= 45 & customer_age < 65 ~ "45to64",
      TRUE ~ "65plus"
    ),
    age_group = factor(age_group)
  )

# 3.3 Check for NA's and simple imputation
summary_df <- ecom_df %>% summarise_all(~ sum(is.na(.)))
print(summary_df)
# If any relevant numeric NA's appear, simple impute (median)
num_vars <- c("price","discount","quantity","delivery_time_days","total_amount","shipping_cost","profit_margin","customer_age","unit_price")
for (v in num_vars) {
  if (any(is.na(ecom_df[[v]]))) {
    median_val <- median(ecom_df[[v]], na.rm=TRUE)
    ecom_df[[v]][is.na(ecom_df[[v]])] <- median_val
  }
}

```
Explanation: convert order_date using ymd(). 
Extract month and weekday as categorical features since return behavior may vary by season/day. 
Create unit_price and high_value_order as potentially predictive features. Group ages. 
Handle missing numeric values via median imputation (simple and acceptable for a project).

4. Exploratory Data Analysis (EDA)

```{r}
library(dplyr)
library(lubridate)

ecom_df <- ecom_df %>%
  mutate(
    # dates
    order_date = ymd(order_date),   # "2023-12-23" → Y-m-d, so use ymd()

    # categorical variables
    category        = factor(category),
    payment_method  = factor(payment_method),
    region          = factor(region),
    returned        = factor(returned),
    customer_gender = factor(customer_gender),

    # numeric variables (safe to re-cast even if already numeric)
    price              = as.numeric(price),
    discount           = as.numeric(discount),
    quantity           = as.numeric(quantity),
    delivery_time_days = as.numeric(delivery_time_days),
    total_amount       = as.numeric(total_amount),
    shipping_cost      = as.numeric(shipping_cost),
    profit_margin      = as.numeric(profit_margin),
    customer_age       = as.numeric(customer_age)
  )

glimpse(ecom_df)

```


```{r}
summary(ecom_df)
```


```{r}
numeric_vars <- c("price","discount","quantity",
"delivery_time_days","total_amount",
"shipping_cost","profit_margin","customer_age")

ecom_df %>%
select(all_of(numeric_vars)) %>%
pivot_longer(cols = everything(), names_to = "variable", values_to = "value") %>%
ggplot(aes(value)) +
geom_histogram(fill = "steelblue", color = "black", bins = 20) +
facet_wrap(~ variable, scales = "free") +
labs(
title = "Distribution of Key Numeric Variables",
x = "Value",
y = "Count"
) +
theme_minimal()

```
```{r}
ecom_df %>%
count(category) %>%
ggplot(aes(x = reorder(category, n), y = n, fill = category)) +
geom_col() +
coord_flip() +
labs(
title = "Order Count by Product Category",
x = "Product Category",
y = "Number of Orders"
) +
theme_minimal()

```
```{r}
# 4.1 Class balance
ecom_df %>% count(returned) %>% mutate(prop = n / sum(n))

# 4.2 Visuals (run in RStudio / RMarkdown to display)
# Returned by category
ecom_df %>% 
  group_by(category, returned) %>% 
  summarise(n = n()) %>% 
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = category, y = prop, fill = returned)) +
  geom_col(position = "dodge") +
  labs(title = "Return rate by product category", y = "Proportion")

# Returned by high_value_order
ecom_df %>% count(high_value_order, returned) %>%
  group_by(high_value_order) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = high_value_order, y = prop, fill = returned)) +
  geom_col(position = "dodge") +
  labs(title = "Return proportion by high value order flag", y = "Proportion")

```
Explanation: check class imbalance and simple visual relationships to guide modeling. 
Looking for features strongly associated with returns (category, high-value orders, shipping time).

The visualization shows the proportion of returned vs. non-returned orders across product categories. The bars indicate that:

Most categories have a very low return rate (Only a small teal portion).

Grocery products show almost zero returns, likely because these items are perishable or low-value.

Fashion has one of the highest return rates, which aligns with real e-commerce behavior—customers often return clothing due to sizing, fit, or style issues.

Electronics and Sports categories also show slightly higher return proportions compared to Beauty, Home, or Grocery.

Overall, returns make up a very small fraction of total orders, indicating that the majority of transactions are successful.

Return rate by customer age group:

```{r}
library(dplyr)
library(ggplot2)

ecom_df %>%
  mutate(age_group = cut(customer_age, 
                         breaks = c(18, 30, 45, 60, 80),
                         labels = c("18–30", "31–45", "46–60", "61+"))) %>%
  group_by(age_group, returned) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(age_group) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = age_group, y = prop, fill = returned)) +
  geom_col(position = "fill") +
  labs(title = "Return Rate by Customer Age Group",
       x = "Age Group",
       y = "Proportion") +
  scale_fill_brewer(palette = "Set2")

```
Average delivery time for returned vs non-returned items

```{r}
ecom_df %>%
  group_by(returned) %>%
  summarise(avg_delivery = mean(delivery_time_days, na.rm = TRUE)) %>%
  ggplot(aes(x = returned, y = avg_delivery, fill = returned)) +
  geom_col() +
  labs(title = "Average Delivery Time by Return Status",
       x = "Returned?",
       y = "Average Delivery Time (days)")

```
Price vs Return Status

```{r}
ggplot(ecom_df, aes(x = returned, y = price, fill = returned)) +
  geom_boxplot() +
  labs(title = "Price Difference Between Returned and Non-Returned Orders",
       x = "Returned",
       y = "Price")

```
5. Train/test split and preprocessing recipe

```{r}
set.seed(1234)
train_index <- createDataPartition(ecom_df$returned, p = 0.75, list = FALSE)
train <- ecom_df[train_index, ]
test  <- ecom_df[-train_index, ]

# recipe: simple preprocessing for modeling
rec <- recipe(returned ~ price + discount + quantity + delivery_time_days +
                shipping_cost + profit_margin + customer_age + unit_price +
                category + payment_method + region + order_month + order_wday +
                high_value_order + age_group,
              data = train) %>%
  step_dummy(all_nominal_predictors(), -all_outcomes()) %>%  # one-hot encoding
  step_zv(all_predictors()) %>%                             # remove zero variance
  step_center(all_numeric_predictors()) %>%
  step_scale(all_numeric_predictors())
rec_prep <- prep(rec, training = train)
x_train <- bake(rec_prep, new_data = train)
x_test  <- bake(rec_prep, new_data = test)

```

Explanation: split 75/25. 
Using a recipe to one-hot encode categorical predictors and standardize numeric predictors; 
preparing data to feed to models that require numeric input. 
step_dummy() makes factors suitable for glm and xgboost

6. Baseline model: Logistic Regression (optional quick baseline)

```{r}
set.seed(1234)
glm_fit <- train(returned ~ ., data = x_train,
                 method = "glm",
                 family = "binomial",
                 trControl = trainControl(method = "cv", number = 5, classProbs = TRUE, summaryFunction = twoClassSummary),
                 metric = "ROC")

glm_fit
glm_pred <- predict(glm_fit, newdata = x_test)
glm_probs <- predict(glm_fit, newdata = x_test, type = "prob")[, "Yes"]

confusionMatrix(glm_pred, x_test$returned, positive = "Yes")
glm_roc <- roc(response = x_test$returned, predictor = glm_probs, levels = c("No","Yes"))
auc(glm_roc)

```
Explanation: building logistic model as a baseline; 
reporting confusion matrix and AUC.

Interpretation of Baseline Logistic Regression Model
1. Overall Model Performance

Your logistic regression model was trained on 25,876 samples, with 44 predictors and a binary target variable (Yes = returned, No = not returned). The model used 5-fold cross-validation, which is a good practice for baseline performance estimation.

The model produced an ROC = 0.5887, which indicates:

The model is only slightly better than random guessing (AUC = 0.5).

It struggles to separate returned vs. non-returned orders.

This is expected when predicting rare events like product returns.

AUC below 0.60 typically means the features may not have strong linear relationships with the target, or class imbalance is impacting performance.

2. Class Imbalance and Prediction Behavior

```{r}
#Prediction   No   Yes
#No          8149  475
#Yes            0    0

```

This means:

The model never predicted “Yes” (returned) for any observation.

It predicted every case as “No” (not returned).

This is a classic sign of severe class imbalance — the return rate in your dataset is around 5.5%, which is very low. Logistic regression, without rebalancing, tends to default to the majority class because it maximizes accuracy by ignoring rare outcomes.

This explains why:

Sensitivity (Recall for Yes) = 0.00

Specificity (Recall for No) = 1.00

Balanced Accuracy = 0.50

Kappa = 0 (no better than random agreement)

Even though the model shows a high accuracy = 94.49%, this number is misleading because simply predicting “No” for everyone gives the same accuracy.

Conclusion: Logistic regression baseline has zero ability to detect returned items.

3. No Information Rate (NIR) Insight

```{r}
# No Information Rate : 0.9449
# P-Value [Acc > NIR] : 0.5122

```

NIR = accuracy of always predicting the majority class (“No”), which is 94.49%.

Your model’s accuracy = exactly the same as NIR.

P-value shows the model is not significantly better than the trivial baseline.

This is further proof the model is not learning meaningful patterns in its current form.

4. AUC (ROC = 0.6236): What It Means

Even though the confusion matrix is poor, ROC = 0.6236 suggests:

The model does rank returned items as slightly more “return-prone” than non-returned ones.

But the default probability threshold (0.5) hides that information.

With rare events, the threshold usually needs to be lowered (e.g., 0.1 or 0.2).

This means the logistic model contains some predictive value — just not at the default decision threshold.

5. Why Logistic Regression Failed

Severe class imbalance (only 5% returned)

Default probability threshold too high

Linear model unable to capture complex relationships

Some predictors may have low variance or minimal correlation with returns

This makes logistic regression a weak baseline for this dataset.




Fix Class Imbalance

A. SMOTE (Best starting point)

```{r}
install.packages("ROSE")   # if not installed
library(ROSE)

set.seed(123)
# Split first
train_index <- createDataPartition(ecom_df$returned, p = 0.75, list = FALSE)
train <- ecom_df[train_index, ]
test  <- ecom_df[-train_index, ]

# Use ovun.sample to create a balanced sample
rose_out <- ovun.sample(returned ~ ., data = train, method = "both", p = 0.5, N = nrow(train), seed = 1)$data
table(rose_out$returned)

# Train model on rose_out
ctrl <- trainControl(method = "cv", number = 5, classProbs = TRUE, summaryFunction = twoClassSummary)
set.seed(123)
xgb_rose <- train(returned ~ ., data = rose_out, method = "xgbTree", metric = "ROC", trControl = ctrl)
xgb_rose

# Evaluate
probs_rose <- predict(xgb_rose, newdata = test, type = "prob")[, "Yes"]
preds_rose <- predict(xgb_rose, newdata = test)
confusionMatrix(preds_rose, test$returned, positive = "Yes")


```


7. Method 1 (Chapter 1–10): Decision Tree (rpart)

```{r}
# Train a decision tree with caret wrapper
ctrl <- trainControl(method = "cv", number = 5, classProbs = TRUE, summaryFunction = twoClassSummary)

set.seed(1234)
rpart_grid <- expand.grid(cp = seq(0.001, 0.05, length.out = 8))

rpart_fit <- train(returned ~ ., data = x_train,
                   method = "rpart",
                   trControl = ctrl,
                   tuneGrid = rpart_grid,
                   metric = "ROC")

rpart_fit
# plot the final tree
rpart.plot(rpart_fit$finalModel, main = "Decision tree for returned")

# Predictions and evaluation
rpart_pred <- predict(rpart_fit, newdata = x_test)
rpart_probs <- predict(rpart_fit, newdata = x_test, type = "prob")[, "Yes"]
confusionMatrix(rpart_pred, x_test$returned, positive = "Yes")
rpart_roc <- roc(response = x_test$returned, predictor = rpart_probs, levels = c("No","Yes"))
auc(rpart_roc)

```
Explanation: training a CART decision tree (interpretable) with 5-fold CV tuning complexity parameter cp. 
Evaluating with confusion matrix and ROC AUC. 
Using rpart.plot to inspect splits

Just like the logistic regression baseline, the decision tree never predicts the minority class (“Yes”) even once. 
It predicts “No” for all observations.

8. Method 2 (Chapter 11–15): Gradient Boosting (xgboost via caret)

```{r}
# Convert data to matrix and label for xgboost
# caret can handle xgboost; create tuning grid
xgb_grid <- expand.grid(nrounds = c(100, 200),
                        max_depth = c(3,6),
                        eta = c(0.05, 0.1),
                        gamma = 0,
                        colsample_bytree = 0.8,
                        min_child_weight = 1,
                        subsample = 0.8)

ctrl_xgb <- trainControl(method = "cv", number = 5, classProbs = TRUE, summaryFunction = twoClassSummary, verboseIter = TRUE)

set.seed(1234)
xgb_fit <- train(returned ~ ., data = x_train,
                 method = "xgbTree",
                 trControl = ctrl_xgb,
                 tuneGrid = xgb_grid,
                 metric = "ROC",
                 verbose = FALSE)

xgb_fit
# Variable importance
vip(xgb_fit$finalModel)

# Predictions & evaluation
xgb_pred <- predict(xgb_fit, newdata = x_test)
xgb_probs <- predict(xgb_fit, newdata = x_test, type = "prob")[, "Yes"]
confusionMatrix(xgb_pred, x_test$returned, positive = "Yes")
xgb_roc <- roc(response = x_test$returned, predictor = xgb_probs, levels = c("No","Yes"))
auc(xgb_roc)

```
Explanation: training boosted trees with caret's xgbTree wrapper. 
Tuning nrounds, max_depth, and eta. 
Report AUC and confusion matrix. Use vip to show feature importance for interpretation.

9. Model comparison & selection

```{r}
results <- resamples(list(Logistic = glm_fit, Tree = rpart_fit, XGBoost = xgb_fit))
summary(results)

# Compare test set ROC AUCs we computed
tibble(
  model = c("Logistic", "DecisionTree", "XGBoost"),
  auc = c(auc(glm_roc), auc(rpart_roc), auc(xgb_roc))
) %>% arrange(desc(auc))

```

Explanation: comparing cross-validated resampling results and hold-out AUC on the test set. 
WE prefer the model with best hold-out AUC and acceptable confusion matrix tradeoffs.

10. Interpret results & business actions

```{r}
# Example: create predictions on entire dataset to flag high-risk orders
x_all <- bake(rec_prep, new_data = ecom_df)
final_probs <- predict(xgb_fit, newdata = x_all, type = "prob")[, "Yes"]
ecom_df$pred_return_prob <- final_probs
ecom_df$pred_return_flag <- ifelse(final_probs > 0.5, "Yes", "No")

```

Explanation: using final model (e.g., xgboost) to score orders. 
Flagging those with predicted return probability above threshold (0.5 here; 
business can tune threshold to target specific precision/recall tradeoff).

Conclusions — summary

