---
title: "Text_ML_Data622"
author: "Mubashira Qari"
date: "2025-11-02"
output: html_document
---

```{r}
# Install packages if not already installed
packages <- c("dplyr", "tidyr", "stringr", "tidytext",
              "tm", "SnowballC", "text2vec",
              "Matrix", "caret", "e1071", "glmnet", "ggplot2", "pROC")
new_pkgs <- packages[!(packages %in% installed.packages()[,"Package"])]
if(length(new_pkgs)) install.packages(new_pkgs)

library(dplyr)
library(tidyr)
library(stringr)
library(tidytext)
library(tm)
library(SnowballC)
library(text2vec)
library(Matrix)
library(caret)
library(e1071)   # NaiveBayes + SVM (caret also wraps these)
library(ggplot2)
library(pROC)

```

tidytext: tokenization, stopwords, etc. using tidyverse style

tm / SnowballC: classic text cleaning & stemming

text2vec: builds fast sparse TF-IDF matrices, good for ML

caret: splits data, trains models, evaluates

PART 1. Example data

```{r}

incidents_df <- tibble(
  sys_id = c("INC001","INC002","INC003","INC004","INC005","INC006"),
  short_description = c(
    "VPN access failing for remote user after password reset",
    "Email server down, cannot send or receive messages company-wide",
    "Database connection timeout on finance reporting dashboard",
    "Laptop blue screen when launching Teams video call",
    "Network latency and packet loss in APAC region office WAN link",
    "Access denied to shared drive after role change"
  ),
  category = c(
    "Access",
    "Messaging",
    "Database",
    "Desktop",
    "Network",
    "Access"
  )
)

```


```{r}
incidents_big <- incidents_df %>%
  slice(rep(1:n(), each = 100))  # repeat each row 100 times

```

Goal: learn to predict category from short_description.

PART 2. Tokenization, cleaning, normalization

Tokenization = split text into words.
Stopword removal = drop common words that don’t carry meaning (“the”, “and”, “to”).
Lowercasing, punctuation removal, numbers removal = standard text cleanup.
Stemming/lemmatization = reduce words to base form: “failing”, “failed”, “fails” → “fail”.

We’ll do this using tidytext first because it's very clear.

```{r}
# 2.1 basic normalization: lowercase, remove non-letter junk
clean_incidents <- incidents_df %>%
  mutate(short_description_clean = short_description %>%
           str_to_lower() %>%                               # lowercase
           str_replace_all("[^a-z0-9 ]", " ") %>%           # remove punctuation/symbols
           str_replace_all("\\s+", " ") %>%                 # collapse multiple spaces
           str_trim()
  )

clean_incidents

```

Explanation:

str_to_lower(): models shouldn’t treat "VPN" vs "vpn" as different tokens, unless you care about capitalization.

str_replace_all("[^a-z0-9 ]"," "): remove punctuation, slashes, commas, etc. We keep numbers because “401 error” might be important.

Trim and collapse spaces so we don’t end up with "" tokens.

Now: tokenize to 1-word tokens (unigrams), remove stopwords, stem.

```{r}
data("stop_words")  # from tidytext

tokens_df <- clean_incidents %>%
  select(sys_id, category, short_description_clean) %>%
  unnest_tokens(
    word,
    short_description_clean,
    token = "words" # splits on space/punct
  ) %>%
  anti_join(stop_words, by = "word") %>%                 # remove stopwords
  mutate(word = wordStem(word, language = "en")) %>%     # stemming using Snowball
  filter(str_detect(word, "[a-z]"))                      # drop tokens that aren't alphabetic

tokens_df

```

What we have now:

One row per (incident, word)

Common junk removed

“failing” and “failed” both become “fail”

Stopwords like “for”, “after”, “and” removed

If you prefer lemmatization (true dictionary base forms like “denied” → “deny”), R doesn’t have a great base lemmatizer in base packages. Stemming is usually fine at this stage.

PART 3. Create Document-Term Matrix (DTM)

We need to convert text into numeric features ML can understand.
We’ll build:

Bag of Words: counts of each word per ticket

TF-IDF: re-weights rare but important words higher

First: get term counts per incident.

```{r}
term_counts <- tokens_df %>%
  count(sys_id, word) %>%     # how many times each word appears in each ticket
  ungroup()

term_counts

```

Now pivot from long format to wide (DTM: each row = ticket, each column = token):

```{r}

dtm_counts <- term_counts %>%
  pivot_wider(
    names_from = word,
    values_from = n,
    values_fill = list(n = 0)
  )

dtm_counts

```
But this is still a regular data frame. For ML, especially SVM/Naive Bayes, we usually want:

Sparse matrix (efficient for thousands of tokens)

TF-IDF weighting (term importance)

We’ll now compute TF-IDF manually using tidytext, then convert to sparse.

```{r}

tf_idf_df <- term_counts %>%
  bind_tf_idf(term = word, document = sys_id, n = n)

tf_idf_df

```
This gives you:

tf: term frequency within a document

idf: inverse document frequency (how rare across all docs)

tf_idf: tf * idf (final weight)

Now pivot TF-IDF into matrix form:

```{r}

dtm_tfidf <- tf_idf_df %>%
  select(sys_id, word, tf_idf) %>%
  pivot_wider(
    names_from = word,
    values_from = tf_idf,
    values_fill = list(tf_idf = 0)
  )

dtm_tfidf

```
Last: join the label (category) back in.

```{r}
model_df <- clean_incidents %>%
  select(sys_id, category) %>%
  left_join(dtm_tfidf, by = "sys_id")

model_df

```

At this point:

model_df$category is the target

All other columns are numeric TF-IDF features

We are now ready to train models.

PART 4. Split train/test

Important so we can measure generalization.

```{r}
library(caret)

set.seed(123)
model_df$category <- factor(model_df$category)

train_index <- createDataPartition(model_df$category, p = 0.8, list = FALSE)
train_df <- model_df[train_index, ]
test_df  <- model_df[-train_index, ]

x_train <- train_df %>% select(-sys_id, -category)
y_train <- train_df$category

x_test  <- test_df %>% select(-sys_id, -category)
y_test  <- test_df$category

```


```{r}
set.seed(123)

# caret wants the target as factor
model_df$category <- factor(model_df$category)

# create train/test index (80/20 split)
train_index <- createDataPartition(model_df$category, p = 0.8, list = FALSE)

train_df <- model_df[train_index, ]
test_df  <- model_df[-train_index, ]

# Separate predictors and labels
x_train <- train_df %>% select(-sys_id, -category)
y_train <- train_df$category

x_test  <- test_df %>% select(-sys_id, -category)
y_test  <- test_df$category

```
Why?

We don’t train and test on the same data. Otherwise we’d just be memorizing.

Note: With only 6 rows of toy data, the split won’t be stable. In real ServiceNow data you might have thousands of incidents, which is perfect.

PART 5. Train Model 1: Naive Bayes

Why Naive Bayes works well for text:

It assumes each word contributes independently to the class.

It’s fast and often surprisingly strong for text classification.

We’ll let caret train the model with method = "naive_bayes".

```{r}
# --- Load / Install Required Packages ---
if (!require(caret)) install.packages("caret", dependencies = TRUE)
if (!require(e1071)) install.packages("e1071")     # caret depends on this for Naive Bayes
if (!require(MLmetrics)) install.packages("MLmetrics")

library(caret)
library(e1071)
library(MLmetrics)
library(doParallel)

```


```{r}
# Remove zero-variance predictors first
nzv <- nearZeroVar(x_train)
x_train <- x_train[, -nzv]
x_test  <- x_test[, -nzv]

# Ensure serial execution (avoid parallel errors)
library(doParallel)
registerDoSEQ()

ctrl <- trainControl(
  method = "cv",
  number = 5,  # now safe because we have enough data
  classProbs = TRUE,
  summaryFunction = multiClassSummary
)

# Disable kernel density; use Laplace smoothing
nb_grid <- expand.grid(
  usekernel = FALSE,
  laplace = 1,
  adjust = 1
)

nb_model <- train(
  x = x_train,
  y = y_train,
  method = "naive_bayes",
  trControl = ctrl,
  tuneGrid = nb_grid
)

nb_model

```
What’s happening:

caret automatically handles tuning the naive Bayes parameters

cross-validation means: split training set into folds, train on 4/5, validate on 1/5, rotate. This protects you from overfitting.

Evaluate on test set:

```{r}
nb_pred <- predict(nb_model, newdata = x_test)
confusionMatrix(nb_pred, y_test)

```

Interpretation:

Confusion matrix tells you which categories you're predicting correctly vs confusing.

For imbalanced data (e.g., “Access” is very common), look at per-class recall.

PART 6. Train Model 2: SVM

Why SVM for text:

Works well in high-dimensional sparse spaces (exactly what TF-IDF creates).

Often competitive with deep learning for short text like ticket titles.

```{r}
svm_model <- train(
  x = x_train,
  y = y_train,
  method = "svmLinear",   # linear SVM is typical for text
  trControl = ctrl
)

svm_model

```

Predict + evaluate:

```{r}
svm_pred <- predict(svm_model, newdata = x_test)
confusionMatrix(svm_pred, y_test)

```

What to compare:

Overall Accuracy (but careful if imbalanced)

Per-class Recall (did we learn rare classes?)

Kappa (agreement beyond chance)

PART 7. (Optional) Probability curves / ROC

If you have a binary classification (“Major Incident” yes/no or “Security Incident” yes/no), you can compare ROC and AUC:

```{r}
# This only works if you have exactly 2 classes in y_test
if (nlevels(y_test) == 2) {
  svm_probs <- predict(svm_model, newdata = x_test, type = "prob")[,2]
  roc_obj <- roc(response = y_test, predictor = svm_probs)
  plot(roc_obj, main = "SVM ROC Curve")
  auc(roc_obj)
}

```

For multiclass, you can get per-class ROC by treating one class as “positive” vs all others.

PART 8. Scaling up using text2vec (production style)

The tidytext path we walked is super clear to learn.
In production (thousands+ tickets), use text2vec because it:

Streams text (doesn’t need to build huge wide data frame in memory first)

Builds a sparse matrix directly

Fits naturally with machine learning algorithms that expect matrices

High-level text2vec workflow:

Prepare vectorizer (tokenization rules, stopword removal, n-grams)

Build vocabulary

Create document-term matrix

Transform to TF-IDF

Train model on sparse Matrix

Sketch:

```{r}
# 8.1 Prepare the text (lowercase etc.)
text_vec <- clean_incidents$short_description_clean

# labels
labels_vec <- factor(clean_incidents$category)

# 8.2 Tokenizer function
tok_fun <- word_tokenizer  # provided by text2vec

it <- itoken(
  text_vec,
  tokenizer = tok_fun,
  progressbar = FALSE
)

# 8.3 Build vocabulary and prune stopwords
vocab <- create_vocabulary(it, stopwords = stop_words$word) %>%
  prune_vocabulary(term_count_min = 2)  # keep terms that appear >=2 times

# 8.4 Vectorizer
vectorizer <- vocab_vectorizer(vocab)

# 8.5 Build Document-Term Matrix (counts)
dtm <- create_dtm(it, vectorizer)

# 8.6 TF-IDF transform
tfidf_transformer <- TfIdf$new()
dtm_tfidf_mat <- tfidf_transformer$fit_transform(dtm)
# NOTE: dtm_tfidf_mat is a sparse Matrix::dgCMatrix

# 8.7 Train linear model on sparse matrix
svm_model_sparse <- train(
  x = as.matrix(dtm_tfidf_mat),
  y = labels_vec,
  method = "svmLinear",
  trControl = ctrl
)
svm_model_sparse

```

This version is what you'd actually use at scale:

Works with 10k, 100k+ tickets

No manual pivot_wider

Efficient memory usage

RECAP / MENTAL MODEL

Here’s the “story” you should remember for text ML:

Step 1. Collect text + label
short_description → "VPN access failing..."
category → "Access"

Step 2. Clean text
lowercase, strip punctuation, remove stopwords like "the", stem words to base

Step 3. Tokenize
turn each ticket into bag of words:
["vpn","access","fail","remot","password","reset",...]

Step 4. Represent as numbers
use TF-IDF: high weight for words that are rare overall but important in that ticket
Result: big sparse matrix of numbers

Step 5. Split into train/test
never evaluate on training data only

Step 6. Train classifier
Naive Bayes (probabilistic, fast)
SVM (large-margin, strong for text)

Step 7. Evaluate
confusion matrix, recall, AUC (if binary)

Step 8. Improve

Add bigrams ("password reset", "blue screen")

Handle class imbalance (SMOTE)

Remove noisy tickets like "test please ignore"

ACTION FOR YOU

You can paste the full pipeline (Parts 0–7) into RStudio and run it with your incident export (sys_id, short_description, category or assignment_group). The only thing you’d change is: load your real CSV instead of the toy example.

When you’re ready, I can:

Modify this pipeline to work when you do NOT have a category yet (i.e. use clustering to discover topics instead of classification).

Add SMOTE for imbalanced categories.

Add bigrams/trigrams like "password reset", "email server".

